\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2016-12-21 mSystems/rev/mSys_MandM_rev.tex
new study in the text, several minor modifications

%%
\subsubsection*{Sample selection}
We have chosen studies about relevant pathologies containing metagenomic sequencing time data series of bacterial populations from humans in different healthy and non-healthy states. We have selected only those individuals who had three or more time points of data available in databases. Caporaso et al. study \cite{moving} was selected as it has two healthy individuals measured over a very large timespan, with almost daily sampling. Faith et al. study \cite{LEA} was selected due to the BMI differences between subjects, moreover some of them had diets which can be treated as system perturbations. We considered healthy only those individuals who had normal or overweight BMI. Smith et al., study \cite{kwashiorkor} was selected for both the age of the patients and the rare disease. We only worked with the discordant twins, and considered healthy those who were not affected by kwashiorkor in each pair of patients. David et al., study \cite{diet} was selected for its differential diets. We considered as the healthy part the first time samples of each individual before the diet, and the rest of time points as perturbations. Dethlefsen and Relman work \cite{antibiotic} was selected due to the interesting treatment of two antibiotic intakes of the same antibiotic by three different subjects. We considered as the healthy part only those times before any antibiotic treatment, and as perturbations the time of antibiotic intakes and the period after that. And finally, we also considered a study made in our group carried by Durban et al., \cite{IBS} in which we considered as healthy subjects those who didn't suffer from irritable bowel disease, and as perturbation the patients who had this disease.

Metadata of each study is provided in Supplementary Tables S1 to S4. All used 16S rRNA gene sequencing except for the study of the discordant kwashiorkor twins\cite{kwashiorkor}, where both SMS and 16S rRNA data were used. In the latter case we selected to work with SMS data to show that our method is valid regardless of the source of taxonomic information. Each one of the datasets was treated as follows:

>>>>

%%
\subsubsection*{Sample selection}
We have chosen studies about relevant pathologies containing metagenomic sequencing time data series of bacterial populations from humans in different healthy and non-healthy states. We have selected only those individuals who had three or more time points of data available in databases. Caporaso et al. study \cite{moving} was selected as it has two healthy individuals measured over a very large timespan, with almost daily sampling. Faith et al. study \cite{LEA} was selected due to the BMI differences between subjects, moreover some of them had diets which can be treated as system perturbations. We considered healthy only those individuals who had normal or overweight BMI. Smith et al., study \cite{kwashiorkor} was selected for both the age of the patients and the rare disease. We only worked with the discordant twins, and considered healthy those who were not affected by kwashiorkor in each pair of patients. David et al., study \cite{diet} was selected for its differential diets. We considered as the healthy part the first time samples of each individual before the diet, and the rest of time points as perturbations. Dethlefsen and Relman work \cite{antibiotic} was selected due to the interesting treatment of two antibiotic intakes of the same antibiotic by three different subjects. We considered as the healthy part only those times before any antibiotic treatment, and as perturbations the time of antibiotic intakes and the period after that. David et al. work \cite{hostlife} was selected due to the great longitudinal data that it provides plus its complete metadata and the interesting events that happened to both subjects (and infection and a travel). We considered as a healthy part the previous time points before each event. And finally, we also considered a study made in our group carried by Durban et al., \cite{IBS} in which we considered as healthy subjects those who didn't suffer from irritable bowel disease, and as perturbation the patients who had this disease.

Metadata of each study is provided in Supplementary Tables S1 to S7. All used 16S rRNA gene sequencing except for the study of the discordant kwashiorkor twins\cite{kwashiorkor}, where both SMS and 16S rRNA data were used. In the latter case we selected to work with SMS data to show that our method is valid regardless of the source of taxonomic information. Each one of the datasets was treated as follows:

### 2016-12-21 mSystems/rev/mSys_results_rev.tex
new study in the text, several minor modifications

Taylor parameters describing the temporal variability of the gut microbiome in our sampled individuals are shown in Supplementary Tables S1 to S6. Our results hint at an ubiquitous behavior. On the first hand, the variability (which corresponds to the maximum amplitude of fluctuations) is large, which suggests resilient capacity of the microbiota. On the other hand, the scaling index is always smaller than one, which means that more abundant taxa are less volatile than less abundant ones. In addition, Taylor parameters for the microbiome of healthy individuals in different studies are compatible within estimated errors. This enables us to define an area in the Taylor parameter space that we called the \emph{healthy zone}.

In order to jointly visualize and compare the results of individuals from different studies\cite{IBS,moving,antibiotic,LEA,kwashiorkor,diet}, their Taylor parameters have been standardized, where standardization means that each parameter is subtracted by the mean value and divided by the standard deviation of the group of healthy individuals for each study (for details of the procedure, please see Standardization subsection in Material and Methods). The healthy zone and the standardized Taylor parameters for individuals whose gut microbiota is compromised (i.e., suffering from kwashiorkor, altered diet, antibiotics or IBS) is shown in Figure \ref{fig:main2}. Children developing kwashiorkor show smaller variability than their healthy twins. A meat/fish-based diet increases the variability significantly when compared to a plant-based diet. All other cases presented increased variability, which is particularly severe, and statistically significant at more than 95\% CL, for obese patients grade III on a diet, individuals taking antibiotics or IBS--diagnosed patients. A global property emerges from all worldwide data collected: Taylor parameters characterize the statistical behavior of microbiome changes. Furthermore, we have verified that our conclusions are robust to systematic errors due to taxonomic assignment (see Taxa level selection in Material and Methods).

>>>>

Taylor parameters describing the temporal variability of the gut microbiome in our sampled individuals are shown in Supplementary Tables S1 to S7. Our results hint at an ubiquitous behavior. On the first hand, the variability (which corresponds to the maximum amplitude of fluctuations) is large, which suggests resilient capacity of the microbiota. On the other hand, the scaling index is always smaller than one, which means that more abundant taxa are less volatile than less abundant ones. In addition, Taylor parameters for the microbiome of healthy individuals in different studies are compatible within estimated errors. This enables us to define an area in the Taylor parameter space that we called the \emph{healthy zone}.

In order to jointly visualize and compare the results of individuals from different studies\cite{IBS,moving,antibiotic,LEA,kwashiorkor,diet,hostlife}, their Taylor parameters have been standardized, where standardization means that each parameter is subtracted by the mean value and divided by the standard deviation of the group of healthy individuals for each study (for details of the procedure, please see Standardization subsection in Material and Methods). The healthy zone and the standardized Taylor parameters for individuals whose gut microbiota is compromised (i.e., suffering from kwashiorkor, a Salmonella infection, travel abroad, altered diet, antibiotics or IBS) is shown in Figure \ref{fig:main2}. Children developing kwashiorkor show smaller variability than their healthy twins. A meat/fish-based diet increases the variability significantly when compared to a plant-based diet. All other cases presented increased variability, which is particularly severe, and statistically significant at more than 95\% CL, for obese patients grade III on a diet, individuals taking antibiotics, the subject who had a Salmonella infection, the subject who did a travel abroad or the IBS--diagnosed patients. A global property emerges from all worldwide data collected: Taylor parameters characterize the statistical behavior of microbiome changes. Furthermore, we have verified that our conclusions are robust to systematic errors due to taxonomic assignment (see Taxa level selection in Material and Methods).
